Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/interactive.tex
compact fac-steps to save space
Wes minor changes to the second half of the paper. See WRW comments.

Next, we present a novel way to debug executions.
%
First, we develop a notion of a \emph{reduction graphs}
and extend our semantics with a form of \emph{tracing}
so that they incrementally collect the edges
in the graph (\S~\ref{sec:inter-semant}).
%
Next, we express a set of common \emph{interactive debugging}
steps as graph traversals (\S~\ref{sec:traversing-graph}), yielding
an novel interactive debugger that allows the user
to effectively visualize \emph{how} the program goes (wrong.)

\subsection{Tracing Semantics}
\label{sec:inter-semant}
%
\begin{figure*}[t]
%% \relDescription{Computing Sub-Terms}
%% \begin{gather*}
%% \begin{array}{lcl}
%% \subtermssym                 & \dcolon & e \to \tr \\
%% \subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
%% \subterms{\eplus{e_1}{e_2}}   & \defeq & \subterm{\eplus{e_1}{e_2}}{e_1}; \subterm{\eplus{e_1}{e_2}}{e_2} \\
%% \subterms{\eif{e_1}{e_2}{e_3}}   & \defeq & \subterm{\eif{e_1}{e_2}{e_3}}{e_1}; \\
                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_2}; \\
                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_3} \\
%% \subterms{\elet{x}{e_1}{e_2}}   & \defeq & \subterm{\elet{x}{e_1}{e_2}}{e_1}; \\
                                %% &        & \subterm{\elet{x}{e_1}{e_2}}{e_2} \\
%% \subterms{\efun{x}{e}}       & \defeq & \subterm{\efun{x}{e}}{e} \\
%% \subterms{e}                 & \defeq & \bullet
%% \end{array}
%% \end{gather*}
\judgementHead{Traced Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
\begin{gather*}
\inference[\recontext]
  {\stepg{e}{\su}{\tr}{e_1}{\su_1}{\tr_1}}
  {\stepg{C[e]}{\su}{\tr}{C[e_1]}{\su_1}{\singlestep{C[e]}{C[e_1]}; \subterms{C[e_1]}; \tr_1}}
\\ \\
\inference[\reappgood]
  {\pair{\efun{x}{e}}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}
         {e\sub{x}{v_2}}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \subterms{e\sub{x}{v_2}}; \tr}}
\\ \\
\inference[\reappbad]
  {\pair{\stuck}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}{\stuck}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
\end{gather*}
\caption{A selection of the operational semantics from
  Figure~\ref{fig:operational}, extended to collect a full reduction
  graph.}
\label{fig:interactive}
\end{figure*}

\paragraph{Reduction Graphs}
%
A \emph{steps-to} edge is a pair of expressions \singlestep{e_1}{e_2}, which
intuitively indicates that $e_1$ steps, in a single step, to $e_2$.
%
%A \emph{sub-term} edge is a pair of expressions \subterm{e_1}{e_2}, which
%intuitively indicates that $e_1$ contains $e_2$ as a sub-expression.
%
A \emph{reduction graph} is a set of steps-to edges:
$$\tr ::= \bullet \spmid \singlestep{e}{e}; \tr$$  %% \spmid \subterm{e}{e}; \tr$$

@@

Next, we show how to build a visual interactive debugger
from the traced semantics, by describing the visualization
\emph{state} \ie\ what the user sees at any given moment,
the set of \emph{commands} available to user and what
they do, and finally how we use a command to \emph{update}
the visualization state. In the sequel, for clarity of
exposition, we assume we have a (global) trace:
$\stepg{e_0}{\emptysu}{\bullet}{e_n}{\vsu}{\tr}$, where
$e_0$ and $e_n$ are the \emph{initial} and \emph{final}
expressions respectively.

>>>>

Next, we present a novel way to debug executions.
%
First, we develop a notion of a \emph{reduction graphs}
and extend our semantics with a form of \emph{tracing}
so that they incrementally collect the edges
in the graph (\S~\ref{sec:inter-semant}).
%
Next, we express a set of common \emph{interactive debugging}
steps as graph traversals (\S~\ref{sec:traversing-graph}), yielding
an novel interactive debugger that allows the user
to effectively visualize \emph{how} the program goes (wrong).

\subsection{Tracing Semantics}
\label{sec:inter-semant}
%
\begin{figure*}[t]
%% \relDescription{Computing Sub-Terms}
%% \begin{gather*}
%% \begin{array}{lcl}
%% \subtermssym                 & \dcolon & e \to \tr \\
%% \subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
%% \subterms{\eplus{e_1}{e_2}}   & \defeq & \subterm{\eplus{e_1}{e_2}}{e_1}; \subterm{\eplus{e_1}{e_2}}{e_2} \\
%% \subterms{\eif{e_1}{e_2}{e_3}}   & \defeq & \subterm{\eif{e_1}{e_2}{e_3}}{e_1}; \\
                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_2}; \\
                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_3} \\
%% \subterms{\elet{x}{e_1}{e_2}}   & \defeq & \subterm{\elet{x}{e_1}{e_2}}{e_1}; \\
                                %% &        & \subterm{\elet{x}{e_1}{e_2}}{e_2} \\
%% \subterms{\efun{x}{e}}       & \defeq & \subterm{\efun{x}{e}}{e} \\
%% \subterms{e}                 & \defeq & \bullet
%% \end{array}
%% \end{gather*}
\judgementHead{Traced Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
\begin{gather*}
\inference[\recontext]
  {\stepg{e}{\su}{\tr}{e_1}{\su_1}{\tr_1}}
  {\stepg{C[e]}{\su}{\tr}{C[e_1]}{\su_1}{\singlestep{C[e]}{C[e_1]}; \subterms{C[e_1]}; \tr_1}}
\\ \\
\inference[\reappgood]
  {\pair{\efun{x}{e}}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}
         {e\sub{x}{v_2}}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \subterms{e\sub{x}{v_2}}; \tr}}
\\ \\
\inference[\reappbad]
  {\pair{\stuck}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}{\stuck}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
\end{gather*}
\caption{A selection of the operational semantics from
  Figure~\ref{fig:operational}, extended to collect a full reduction
  % WRW: \subterms is Used here in the figure before being Defined (indeed,
  % it is never defined anywhere that I can see). A simple bit in text to
  % explain what you mean would help. Why is subterms necessary? What does
  % it collect up for you? 
  graph.}
\label{fig:interactive}
\end{figure*}

\paragraph{Reduction Graphs}
%
A \emph{steps-to} edge is a pair of expressions \singlestep{e_1}{e_2}, which
% WRW: This was "e_1 steps, in a single step" which was circular.
intuitively indicates that $e_1$ reduces, in a single step, to $e_2$.
%
%A \emph{sub-term} edge is a pair of expressions \subterm{e_1}{e_2}, which
%intuitively indicates that $e_1$ contains $e_2$ as a sub-expression.
%
A \emph{reduction graph} is a set of steps-to edges:
$$\tr ::= \bullet \spmid \singlestep{e}{e}; \tr$$  %% \spmid \subterm{e}{e}; \tr$$

@@

Next, we show how to build a visual interactive debugger
from the traced semantics, by describing the visualization
\emph{state} \ie\ what the user sees at any given moment,
the set of \emph{commands} available to user and what
they do, and finally how we use a command to \emph{update}
the visualization state. In what follows, for clarity of
exposition, we assume we have a (global) trace:
$\stepg{e_0}{\emptysu}{\bullet}{e_n}{\vsu}{\tr}$, where
$e_0$ and $e_n$ are the \emph{initial} and \emph{final}
expressions respectively.
